---
title: "Assignment 3: Data Exploration"
author: "Cate Jaffe"
date: February 8, 2021
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <2/9/2021>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

```{r setup, message = FALSE}

# check working directory, reset if needed
getwd()

# set working directory for all chunks to home folder
knitr::opts_knit$set(root.dir = "C:/ENV872/Environmental_Data_Analytics_2021")

```

```{r library, message = FALSE}

library(tidyverse)
library(lubridate)

```

```{r import data, message = FALSE}

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", 
                    stringsAsFactors = TRUE)

Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
                   stringsAsFactors = TRUE)

```



## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Insects, inlcuding bees, flies, and butterfiles are major drivers of pollination in cultivated and un-cultivated ecosystems. Given their ubiquitous use in agriculture, the impacts of neonicotinoids on insects that pollinate crops used for food, animal feed, and textiles is of particular concern. Beyond this potential to impact pollinating insects, water-soluble neonicotinoids can spread beyond the aagricultural fields where they are applied. In fact, only 5% of the active neonicotinoid ingredient is typically taken up by crop plants, the rest makes its way to other ecosystems and environments (Wood et al. 2017). Many studies have shown that insect opoulations, particularly bees, can be negatively impacted by the active neonicotinoid ingredients (Klein et al. 2007). This is especially troubling given the wide dispersal potential of the insecticide.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Litter and woody debris provide a critical input of organic matter and nutrients to underlying soils. Soils with higher contects of organic matter are often good facilitators of productive plant growth and biological activity. The rate at which litter is broken down can also shed light on ecosystem function and biological activity in a particular system. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Mass data are measured to an accuracy of .01 grams. Weights < .01 gram are reported to indicate the presence of a functional group potentially identified during sorting, but not present at detectable masses.
*Litter is defined as material that is dropped from the forest canopy and has a butt end diameter <2cm and a length <50 cm. Fine wood debris is defined as
material that is dropped from the forest canopy and has a butt end diameter <2cm and a length >50 cm
*Ground traps are sampled once per year. Target sampling frequency for elevated traps varies by vegetation present at the site: frequent sampling (1x every 2weeks) in deciduous forest sites during senescence, infrequent year-round sampling (1x every 1-2 months) at evergreen sites. Sampling in elevated traps in deciduous forests may be discontinues for up to 6 months in leaf-off conditions.



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}

dim(Neonics)
# Neonics dataset has 4,623 observations (rows) of 30 variables (columns)

```

6. Using the `summary` function on the "Effects" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}

summary(Neonics$Effect)

```

> Answer: The most common effects that are studied are Population (1,803 studies), Mortality (1,493), and Behavior (360). Studying mortality and population are both clear choices, as mortality is the most severe impact the toxins can have on insects. Individual mortailty could result in rapid species decline and ecosystem distruption. The same is true of population studies - a declining insect population can be a concerning indicator of ecosystem function and species health. Changes in insect behavior, while not nessecarily fatal, may have concequences for the efficiency of pollination or reproductive success. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}

summary(Neonics$Species.Common.Name)

```

> Answer: The six most commonly studied insects are: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. These are all either bees or wasps, species groups which have been shown in labratory settings to have lethal and sublethal behavioral reactions to neonicotinoids (Blacquière et al. 2012). About 35% of the worlds crops depends on pollinators, and bees are a crucial and prolific group of such pollinators. As providers of important ecosystem services such as food crop production, it makes sense that bees and wasps would be of particular interest for study.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}

# print class 
class(Neonics$Conc.1..Author.)

# assess individual values within "Conc.1..Author." field.
# unique(Neonics$Conc.1..Author.)

```

> Answer: The class of this field is "factor", not numeric. Even though the values in this field are numeric some values are accompanied by characters such as "/" and "~". Thus, R will read these values as factors and import the entire field as a factor.


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r, message = FALSE}

ggplot(Neonics) + 
  geom_freqpoly( aes(x = Publication.Year), bins = 50, lty = 1)


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r, message = FALSE}

ggplot(Neonics) + 
  geom_freqpoly( aes(x = Publication.Year, color = Test.Location), bins = 50, lty = 1)


```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are "field natural" and "lab". Field natural tests appear to have peaked in popularity just prior to 2010, around 2008 or 2009. Lab based tests steadily gained popularity through the early 2000's, peaking in use around 2013 and 2014. Both methods of testing (and testing in general) appears to have tapered out after 2017.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r, message = FALSE }

ggplot(Neonics) + 
  geom_histogram( aes(x = Endpoint), stat = "count" ) + # graph "count" of Endpoints
  theme(axis.text.x = element_text(angle = 90)) # tilt axis labels to make more readable


```

> Answer: The two most common endpoints are "NOEL" (No Observable Effect Level) and "LOEL" (Lowest Obervable Effect). NOEL is defined as the highest dose concentration that produces effects which are not significantly different that the control responses. LOEL is defined as the lowest dose concentration that produces effects significantly different from control responses (coded as LOEAL or LOEC).

>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}

class(Litter$collectDate)
# class is factor

# change to date
Litter$collectDate <- ymd(Litter$collectDate)

# check class
class(Litter$collectDate)

# display unique dates
unique(Litter$collectDate)

# Litter was sampled on August 8, 2018 and August 30, 2018.

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

unique(Litter$plotID)

```
> Answer: 12 plots were sampled at Niwot Ridge. When applied to an attirbute of type 'factor' the summary() function returns a count of the number of observations of each type within the field. The unique() function only returns the names of unique values in that field, but does not return the number of times each value occurs in the field. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r, message = FALSE}

ggplot(Litter) + 
  geom_bar( aes(x = functionalGroup) )


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r, message = FALSE}

# boxplot
ggplot(Litter) + 
  geom_boxplot( aes(x = functionalGroup, y = dryMass) )


# violin
ggplot(Litter) + 
  geom_violin( aes(x = functionalGroup, y = dryMass) )


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The benefit of the violin plot is that it displays the density distribution of the plotted data by group, while a boxplot does not. However, with these data it appears that many of the observations are clustered around low numbers or near 0, with a few observations at higher values. This makes it hard to see the shape of the density distributions in the violin plot, and the resulting plot doesn't provide much information about density distributions or group means. In the box plot, it is possible to see group means for some of the functional groups. Both the box plot and the violin plot might benefit from a log transformed y access to address the problem of observations clustered around 0.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles.


## Works Cited

Blacquière, T., Smagghe, G., van Gestel, C.A.M. et al. Neonicotinoids in bees: a review on concentrations, side-effects and risk assessment. Ecotoxicology 21, 973–992 (2012). https://doi.org/10.1007/s10646-012-0863-x

Klein, A. M., Vaissiere, B. E., Cane, J. H., Steffan-Dewenter, I., Cunningham, S. A., Kremen, C., & Tscharntke, T. (2007). Importance of pollinators in changing landscapes for world crops. Proceedings of the royal society B: biological sciences, 274(1608), 303-313.

Wood, T. J., & Goulson, D. (2017). The environmental risks of neonicotinoid pesticides: a review of the evidence post 2013. Environmental science and pollution research international, 24(21), 17285–17325. https://doi.org/10.1007/s11356-017-9240-x

