---
title: "Assignment 7: Time Series Analysis"
author: "Cate Jaffe"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on time series analysis. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A07_TimeSeries.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 16 at 11:59 pm.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r setup, message = FALSE}
#1
# check working directory
getwd()


# Library
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(colormap)


mytheme <- theme_classic( base_size = 14) + 
  theme( axis.text = element_text( color = "#222222ff"),
         legend.position = "top",
         # remove legend title
         legend.title = element_blank(),
         # margins (top,right,bottom,left)
         axis.title.x = element_text( color = "black",
                                    margin = margin(20,0,0,0)),
         axis.title.y = element_text( color = "black",
                                    margin = margin(0,20,0,0)))
theme_set(mytheme)

```

```{r data import}
# temporarily set wd to Ozone folder
setwd("./Data/Raw/Ozone_TimeSeries")
# create list of files in Ozone folder
fileList = list.files(pattern = "*.csv")
# apply read.csv() to each file in list, adding all dataframes to single list object
ozoneFiles = lapply(fileList, read.csv)
# bind rows: OK because columns match for all dataframes
ozoneDat <- bind_rows(ozoneFiles)


## reset wd *for knitting*
knitr::opts_knit$set( root.dir = 
                        "C:/ENV872/Environmental_Data_Analytics_2021",
                  tidy.opts = list(width.cutoff = 60),
                  tidy = TRUE)
# reset wd
setwd("C:/ENV872/Environmental_Data_Analytics_2021")
getwd()

## clean up environment
rm(ozoneFiles, fileList)

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3 
# use lubridate to convert Date (currently in mdy format as character type) to Date type
ozoneDat$Date <- mdy(ozoneDat$Date)

# 4
ozoneDat <- ozoneDat %>% select(Date, 
                                Daily.Max.8.hour.Ozone.Concentration, 
                                DAILY_AQI_VALUE)
# 5
Days <- as.data.frame(seq(ymd("2010-01-01"),
                             ymd("2019-12-31"), 
                             by = "day")) %>% 
  # rename weirdly named column....
  rename("Date" = "seq(ymd(\"2010-01-01\"), ymd(\"2019-12-31\"), by = \"day\")")


# 6
GaringerOzone <- left_join(Days, ozoneDat, by = "Date")

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7
plot_ozoneVtime <- 
  ggplot(GaringerOzone, aes(x = Date, 
                            y = Daily.Max.8.hour.Ozone.Concentration )) + 
  # line plot
  geom_line() + 
  # linear trend line
  geom_smooth(method = "lm") +
  # format x axis
  scale_x_date( breaks = "1 year",
                date_labels = "%Y" ) + 
  theme( axis.text.x = element_text(angle = 45,
                                    hjust = 1)) + 
  # format y axis
  ylab( "Daily Max Ozone (ppm)")
  

print(plot_ozoneVtime)
  

```

>Answer: There appears to be a slight negative trend in Ozone concentrations, meaning that Ozone concentrations appear to be decreasing over time between 2010-2019. However, the the graph makes this downwrad trend appear visually minimal.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r clean NAs}
#8
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)

GaringerOzone <- 
  GaringerOzone %>% 
  mutate( ozone_clean = zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration))
#check
summary(GaringerOzone$ozone_clean)


```

> Answer: 


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r monthly summarize}
#9
GaringerOzone.monthly <- 
  GaringerOzone %>% 
  mutate(month_year = mdy(paste(month(Date),"-01-",year(Date)))) %>% 
    group_by(month_year) %>%
    summarise( mean = mean(ozone_clean))
  
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r time series object}
#10
f_month = month(first(GaringerOzone$Date))
f_day = day(first(GaringerOzone$Date))
f_year = year(first(GaringerOzone$Date))

GaringerOzone.daily.ts <- ts(GaringerOzone$ozone_clean,
                             start = c(f_year, f_month, f_day),
                             frequency = 365)

GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean,
                               start = c(f_year, f_month),
                               frequency = 12)  
  
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r decompose}
#11
dailyDecomp <- stl(GaringerOzone.daily.ts, s.window = "periodic")
plot(dailyDecomp)

monthlyDecomp <- stl(GaringerOzone.monthly.ts, s.window = "periodic")
plot(monthlyDecomp)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r trend analysis}
#12
trendMonthly <- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)
trendMonthlySMK <- smk.test(GaringerOzone.monthly.ts)

summary(trendMonthly)
summary(trendMonthlySMK)

```

> Answer: The seasonal component of variation in the monthly data appears to have a cyclic "up and down" pattern in the components plot above. The other monotonic trend analysis options do not assume seasonality, but the Seasonal Mann Kendall does assume seasonality in the data and accomodates non-normal (non-parametric) data as well.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r plot}
# 13
GaringerOzone.monthly %>% 
  ggplot(aes( x = month_year, y = mean)) + 
  geom_point(color = "#0b4a92ff", alpha = .5) + 
  geom_line(color = "#0b4a92ff", alpha = .8) + 
  scale_x_date(date_labels = "%Y", 
               breaks = "1 year") + 
  labs( x = "Date", 
        y = "Mean Ozone Concentration (ppm)") + 
  ggtitle("Mean Monthly Ozone Concentrations between 2010 and 2019") + 
  theme(plot.title = element_text(vjust = 1))



```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: We started with the reasearch question: Have ozone concentrations changed over the 2010s at this station? Thus, our null hypothesis was that ozone concentrations have not changed overtime between 2010 and 2019. Based on this analysis, we can reject the null hypothesis that ozone concentrations have not changed over the 2010's (p = .0467). That is, it appears ozone concentrations have statistically significantly changed in the 2010's. It should be noted this trend is just barely under the significance threshold of p = .05.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15
# extract components to DF 
componentsMonthly <- as.data.frame(monthlyDecomp$time.series)

componentsMonthly <- 
  componentsMonthly %>% 
  mutate(observed = GaringerOzone.monthly$mean,
         month_year = GaringerOzone.monthly$month_year,
         no_season = observed - seasonal)
  
#16
trendMonthlyMK <-
  Kendall::MannKendall(componentsMonthly$no_season)
summary(trendMonthlyMK)

# plot
componentsMonthly %>% 
  ggplot() + 
  geom_line(aes( x = month_year, y = observed), 
             color = "grey", alpha = .5) + 
  geom_line(aes( x = month_year, y = no_season),
            color = "#0b4a92ff", alpha = .8) + 
  scale_x_date(date_labels = "%Y", 
               breaks = "1 year") + 
  labs( x = "Date", 
        y = "Mean Ozone Concentration (ppm)",
        subtitle = "Raw observations shown in gray, non-seaonsal data shown in blue") + 
  ggtitle("Mean Monthly Ozone Concentrations between 2010 and 2019") + 
  theme(plot.title = element_text(vjust = 1),
        plot.subtitle = element_text(size = 10,
                                     color = "grey" )) 

```

> Answer: This test with non-seasonal data also shows that Ozone concentrations were changing overtime between 2010 and 2019 (p < .01). The statistical significance of the trend with the non-seasonal data is stronger (p = .0467 vs. p = .007 for non-sesasonal data).
