---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Cate Jaffe"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.
```{r library, message=FALSE}
# wrangling
library(tidyverse)
library(agricolae)
library(lubridate)

# plotting
library(viridis)
library(RColorBrewer)
library(colormap)
library(ggplot2)


```

```{r setup, message=FALSE}
#1
# set notebook wd for knit
knitr::opts_knit$set( root.dir = "C:/ENV872/Environmental_Data_Analytics_2021",
                  tidy.opts = list(width.cutoff = 60),
                  tidy = TRUE)
# check wd
#getwd()

# data import
lakeDat <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                    stringsAsFactors = TRUE)
# Date attribute to Date
lakeDat$sampledate <- mdy(lakeDat$sampledate)

#2
mytheme <- theme_light( base_size = 14) + 
  theme( axis.text = element_text( color = "#222222ff"),
         legend.position = "top",
         # remove legend title
         legend.title = element_blank(),
         # margins (top,right,bottom,left)
         axis.title.x = element_text( color = "black",
                                    margin = margin(20,0,0,0)),
         axis.title.y = element_text( color = "black",
                                    margin = margin(0,20,0,0)))
theme_set(mytheme)

```

```{r color reference, include = FALSE}

#scales::show_col(colormap(colormap = colormaps$portland, nshades = 16))

#scales::show_col(colormap(colormap = colormaps$freesurface_blue, nshades = 16))

#scales::show_col(colormap(colormap = colormaps$bathymetry, nshades = 16))

#scales::show_col(colormap(colormap = colormaps$greys, nshades = 16))

#scales::show_col(colormap(colormap = colormaps$rainbow , nshades = 16))


```


## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature in July is not different at varying depths.
Ha: The mean lake temperature in July does differ at different depths.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lakeDat.July <- 
lakeDat %>% 
  # extract month from sampledate, subset to July
  filter(month(sampledate) == 7) %>% 
  #select attributes of interest
  select( "lakename", "year4", "daynum", "depth", "temperature_C") %>% 
  # remove NAs (about 1000 observations)
  na.omit() 


#5
lakeDat.July %>% 
  ggplot(aes( x = depth, y = temperature_C)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm", color = "#52a7a2ff") + 
  ylim(0,35) + 
  labs( x = "Depth (meters)", y = "Temperature (C)") + 
  ggtitle("Lake Temperature by Depth in July")


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that in July lake temperature decreases with increasing depth. That is, shallower waters appear to have higher temperatures, and deeper waters appear to have lower temperatures. While this plot applies a linear regression, the underlying distribution of the points suggests temperature  expotentially increases with decreasing depth.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lm.DepthTemp = lm( data = lakeDat.July, 
                   temperature_C ~ depth)

summary(lm.DepthTemp)

plot(lm.DepthTemp)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: In July, lake temperature and depth are appear to be inversely related, meaning that as lake depth increases we would expect water temperature to decrease. The model indicates that for every 1 meter increase in depth we would expect to see about a 1.9 C decrease in temperature (p < .001, df = 9726). The depth model explains approximately 73.87% of the variation in lake temperature.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
lm.TempAIC <- lm( data = lakeDat.July, 
                       temperature_C ~ year4 + daynum + depth)

step(lm.TempAIC)

# all variables are recommended

#10
lm.TempModel = lm( data = lakeDat.July, 
                   temperature_C ~ year4 + daynum + depth)

summary(lm.TempModel)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method recommended using all variables provided (depth, year, day number). The model including all of these variables explains about 74.1% of the varince in lake temperature in July (p < .001, df = 9724), which is only a .14% improvement over the simpler, depth-only model. From this we might interpret that depth has a much stronger influence on lake temperature in July than the day number or year.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
aov.TempByLake <- aov(data = lakeDat.July,
                      temperature_C ~ lakename)
summary(aov.TempByLake)


lm.TempByLake <- lm(data = lakeDat.July,
                       temperature_C ~ lakename)
summary(lm.TempByLake)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference between the mean temperature in different lakes in July (p < .001, df = 9719). However, the lake only explains 3.87% of the variation in water temperature. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.width = 12, fig.height = 12}
#14.
# all together
lakeDat.July %>% 
  ggplot(aes( x = depth, y = temperature_C, color = lakename)) +
  # jitter points to make more readable
  geom_point(alpha = .5, position = position_jitter()) +
  geom_smooth(method = "lm", se = FALSE) + 
  ylim(0,35) + 
  # axis labels and title
  labs( x = "Depth (meters)", y = "Temperature (C)") + 
  ggtitle("Lake Temperature by Depth in July for Multiple Midwestern Lakes") 


### Overlay of multiple lakes is hard to decipher
# There is no way to cleanly plot tukey labels. 
# Apply facet to clean up graph and to try applying tukey labels.

## make data frame with Tukey labels (also created with code in step 15)
# extract and tukey group labels to dataframe with lakename attached
TempByLake.groups <- 
  HSD.test(aov.TempByLake, "lakename", group = TRUE)
tukeyGrps <- rownames_to_column(TempByLake.groups$groups,
                                var = "lakename")
# clean up dataframe
tukeyLabelDF <- tukeyGrps %>% select("lakename", "groups")

# make plot
lakeDat.July %>% 
  ggplot(aes( x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, color = "#666666ff") + 
  ylim(0,35) + 
  # axis labels and title
  labs( x = "Depth (meters)", y = "Temperature (C)") + 
  ggtitle("Lake Temperature by Depth in July for Multiple Midwestern Lakes, labeled with Tukey Groups") + 
  # facet
  facet_wrap(vars(lakename)) + 
  # remove legend, as facets are labeled
  theme(legend.position = "none") + 
  # Add Tukey groups labels to plot using dataframe
  geom_text(data = tukeyLabelDF, 
            aes(label = groups),
            x = 15, y = 30,
            color = "#666666ff")

 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(aov.TempByLake)

# create and print group labels
TempByLake.groups <- 
  HSD.test(aov.TempByLake, "lakename", group = TRUE)
TempByLake.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake has the same mean temperature (statistically speaking) as Paul Lake and Ward Lake. No lake has a mean temperature that is statistically distinct from all other lakes, as all lakes share a letter group with at least one other lake.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a two-sample t-test, which would test the null hypothesis that the Mean Temperature in Peter Lake is equal to the Mean Temperature in Paul Lake. A p-value less than .05 for this test would indicate that the means are not the same, and are statistically distinct.


